Reduced plasticity and variance in physiological rates for terrestrial ectotherm populations under climate change

### 2024-02-08 ms.qmd
Re-organisation for NEE

## **Introduction**

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Evolutionary theory predicts that greater thermal variability in the past should result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016].  Without plasticity or adaptation, high extinction rates are predicted unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates, thereby buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is only supported for specific life-history stages.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms which may have important cascading effects on energy transfer and productivity across different ecosystems.

@@

@@

Thermal variability (i.e., $CV$) and predictability experienced by a population in the past did not explain acclimation capacity or changes in physiological rate variance among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate; @bush2016incorporating; @seebacher2023can]. For example, incorperating phenotypic plasticity and adaptation into species distribution models has been shown to dampen the predicted contraction of distributions in the face of climate change [e.g., @bush2016incorporating]. It addition, it may be expected that reductions in variance impact the 'opportunity for selection' by reducing the strength of selection and/or the genetic variance exposed to selection, impacting a population's capacity to evolve under climate change [@hoffmann2011climate; @urban2023and].

We show that plastic adjustments in physiological rates and changes in their relative variance differ across habitats. In freshwater habitats, acclimation responses will likely be beneficial and this coincides with relatively little reduction in the relative variance in physiological rates providing more raw material for selection to operate. In contrast, acclimation responses of marine organisms are associated with a nearly equal reduction in relative variance in physiological rates, which may trade-off with the beneficial effects that acclimation provides at high temperatures. In terrestrial ectotherms, relative variance reductions are particularly pronounced and this coincides with a general inability to acclimate. Considering acute physiological rates of animals acclimated to high temperatures are generally elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization may not always provide adaptive benefits under climate change, particularly when considering changes in relative variance.

Our findings highlight the potential vulnerability of terrestrial and marine ectotherms to climate change. Our meta-analytic results provide opportunities to parameterise predictive models in ways that capture the percentage changes in relative variances and acclimation responses when predicting species distributions or modelling population resilience in the face of climate warnming. Our results also define the range of responses observed across diverse taxa (e.g., through prediction intervals) that also provide opportunities to incorperate realistic biological variation into the modelling process. Incorperating relavant variation in such change are needed fully appreciate the biological consequences such changes have on ecological and evolutionary dynamics (@bush2016incorporating).

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Interestingly, there were no differences in acclimation or reductions in physiological variance between life-stages across habitat types. Greater plasticity for early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. However, our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature is equivocal with respect to the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species and the degree to which they can be adjusted in response to the environment may increase our ability to make more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations as evidenced by wide prediction intervals. Nonetheless, changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced and possibly more realistic predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variances changes, models could better decouple how different levels of heritability with different total variance impact evolutionary and ecological predictions. Our meta-analysis allows for realistic models to be parameterised and ensure they are better aligned with empirical findings. Nonetheless, many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate variance change energy flow across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across all major ecosystems.

>>>>

## **Main**

Climate change is expected to result in warmer but also more variable thermal environments for many organisms globally [@Suarez2023; @Easterling2000; @Ummenhofer2017]. Evolutionary theory predicts that greater thermal variability in the past should result in strong selection pressures that lead to genetic adaptation and/or the evolution of adaptive phenotypic plasticity -- both of which are considered important for population resilience to contemporary human-induced climate change [@Seebacher2015-wi; @Merila2014; @Chevin2015; @Chevin2017; @Chevin2010; @Nunney2016; @seebacher2023can; @cooke2021one].  Without plasticity or adaptation, high extinction rates are predicted unless organisms can migrate to track suitable habitats [@Cahill2012; @Nunney2016]. Phenotypic plasticity is expected to be the 'first line of defence' against changing climates, thereby buying time for genetic adaptation to take place [i.e., the 'plasticity first hypothesis' - @WestEberhard2003; see also @Lande2009]. Phenotypic plasticity is predicted to evolve when environmental variability is high but predictable and the costs of plasticity are low [@Chevin2017; @Scheiner2020; @Nunney2016; @Dewitt1998; @Reed2010]. Despite this theoretical expectation, empirical support is scant [but see @Leung2020], likely because many organisms can behaviorally adjust micro-habitat selection to offset thermal stress, the costs of plasticity are high [@Chevin2015; @Dewitt1998], or the prediction is only supported for specific life-history stages.

@@

Periods of past climatic change have had disproportionate impacts on some ecosystems over others raising the question of which ecosystems will be most vulnerable to contemporary climate change. Species occupying terrestrial ecosystems are particularly vulnerable given their weak acclimation abilities and greater probability of experiencing thermal extremes that overwhelm physiological homeostasis [@Seebacher2015-wi; @Gunderson2015-hz; @hoffmann2013upper], whereas greater proximity to upper thermal limits in marine systems have questioned such conclusions [@Pinsky2019-fn]. Marine and freshwater ecosystems appear to have greater physiological acclimation capacity [e.g., @Seebacher2015-wi; @Pottier2022]. However, it is unclear if the magnitude of physiological adjustment is sufficient to compensate for potentially negative environmental effects, particularly when temperature interacts with other abiotic changes. Low oxygen availability may be a major factor influencing the vulnerability of aquatic ecosystems, whereas remaining close to thermal limits and greater water loss is expected to be a stronger constraint on physiological processes in terrestrial ectotherms [@Verberk2016]. Given that terrestrial ectotherms are expected to be closer to their thermal limits, an increase in temperature may have a stronger impact on variation in physiological rates within populations compared to aquatic ectotherms which may have important cascading effects on energy transfer and productivity across different ecosystems [@seebacher2023can; @barneche2021warming].

@@

Using a large database on physiolgical rates (@Seebacher2015-wi) for marine, freshwater and terrestrial ectotherms we apply new effect size estimates [@Noble2022-ty] that capture changes in mean physiological rates ($lnRR_{Q_{10}}$), and changes in their relative variance ($lnCVR_{Q_{10}}$), standardised to a 10$^{\circ}$C temperature difference. These standardised effects sizes can be converted to percentage differences allowing us to compare the relative magnitude of rate changes resulting from acclimation and relative variance change at high temperatures. Acclimation responses could be measured by comparing acclimation $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acclimation}}}$) to acute $lnRR_{Q_{10}}$ ($lnRR_{Q_{10_{acute}}}$) from experimental ectotherms acclimated to different temperatures. The difference in $lnRR_{Q_{10_{acclimation}}}$ compared to  $lnRR_{Q_{10_{acute}}}$ measures the extent of acclimation (i.e., plasticity).

@@

Using the ERA5 climate model, we extracted a 30-year period (1950-2022) of either surface air temperature (0.01$^{\circ}$ resolution) for both terrestrial and freshwater taxa, or sea surface temperature for the marine taxa (at 0.25$^{\circ}$ resolution). We used the historical temperature data for each geographical location to calculate a measure of thermal variability and predcitability (see *Methods*). Thermal variability (i.e., $CV$) and predictability experienced by a population in the past did not explain acclimation capacity or changes in physiological rate variance among terrestrial, marine or freshwater populations (@fig-4; @fig-5).

@@

Understanding the interplay between plasticity and genetic adaptation has important implications for predicting a populations resilience to climate change [@urban2023and; @Chevin2010; @hoffmann2011climate; @merila2014climate; @bush2016incorporating; @seebacher2023can; @cooke2021one]. For example, incorperating phenotypic plasticity and adaptation into species distribution models has been shown to dampen the predicted contraction of distributions in the face of climate change [e.g., @bush2016incorporating]. In addition, it may be expected that reductions in variance impact the 'opportunity for selection' by reducing the strength of selection and/or the genetic variance exposed to selection, impacting a population's capacity to evolve under climate change [@hoffmann2011climate; @urban2023and]. Quntifying the degree of plasticity and genetic/phenotypic variation in key physiolgical responses is therefore recognised as being critical to informing projections for organisms threatened by climate change [@cooke2021one].

We show that plastic adjustments in physiological rates and changes in their relative variance differ across habitats. In freshwater habitats, acclimation responses will likely be beneficial and this coincides with relatively little reduction in the relative variance in physiological rates providing more raw material for selection to operate. In contrast, acclimation responses of marine organisms are associated with a nearly equal reduction in relative variance in physiological rates, which may trade-off with the beneficial effects that acclimation provides at high temperatures. In terrestrial ectotherms, relative variance reductions are particularly pronounced and this coincides with a general inability to acclimate. Considering acute physiological rates of animals acclimated to high temperatures are generally still elevated compared to cold-acclimated animals (~7%; $\beta$ = `r model1.s1$b[3]`, 95% CI: `r model1.s1$ci.lb[3]` to `r model1.s1$ci.ub[3]`, p `r p_value(model1.s1$pval[3])`), acclimatization may not always provide adaptive benefits under climate change, particularly when considering changes in relative variance.

Our findings highlight the potential vulnerability of terrestrial and marine ectotherms to climate change. Our meta-analytic results provide opportunities to parameterise predictive models in ways that capture the percentage changes in relative variances and acclimation responses when predicting species distributions or modelling population resilience in the face of climate warnming. Our results also define the range of biological responses observed across diverse taxa (e.g., through prediction intervals) providing opportunities to incorperate realistic biological variation into the modelling process. Incorperating relavant variation in models is needed fully appreciate the biological consequences such changes in variation and plasticity have on ecological and evolutionary dynamics [@bush2016incorporating; @demarche2019incorporating], that have not yet been explored.

### ***Plasticity and variance in physiological rates do not differ between life-stages***
Life-history stages often occupy different ecological niches and exhibit different physiological responses, levels of plasticity, and patterns of mortality in response to temperature. As such, it is becoming increasingly important to understand how climate change will impact different life-history stages so that more realistic predictive models can be developed [@levy2015resolving; @petitgas2013impacts]. Greater plasticity for early life-stages of development is expected to be important to increase resilience to the effects of climate change because early life-stages are often particularly vulnerable periods in development [@martin2015age; @stearns1976life]. Across a diversity of taxa, we show that there were generally similar patterns between early and late life stages -- both in terms of variance changes and the capacity for plasticity. Our results highlight that early life stages are not likely to be protected from the impacts of climate change. Evidence across the literature is equivocal with respect to the extent to which one life-stage is more or less plastic than another [@carter2020life; @moghadam2019heat]. For example, @moghadam2019heat showed that larval *Drosophila* were more plastic in their heat hardening responses compared to adults. In contrast, @carter2020life showed greater thermal plasticity in metabolism for adults but little in pupae of *Onthophagus taurus*.  Unsurprisingly, there is much variation in the literature and our findings suggest that patterns will depend on the species and traits in question. However, very few studies measure multiple life-history stages within the same population. In future research, there needs to be more studies that explicitly compare differences in plasticity among life stages across multiple traits within the same population.

@@

Enhanced knowledge of how variation in physiological rates vary across populations and species and the degree to which they can be adjusted in response to the environment may increase our ability to make more informed predictions about the ecological and evolutionary dynamics of natural populations [@forsman2015rethinking; @seebacher2023can; @cooke2021one]. While we show general patterns across taxa and habitats it is important to recognise that this does not mean such patterns apply to all populations. Substantial variation in acclimation responses and changes in variance exist among populations as evidenced by wide prediction intervals. Nonetheless, changes in the relative variance in physiological rates could be better incorporated into physiological and ecological models to provide more nuanced and possibly more realistic predictions about the impacts of climate change on natural populations. While we do not yet understand the relative contribution of environmental and genetic factors to variances changes, models could better decouple how different levels of heritability with different total variance impact evolutionary and ecological predictions. Our meta-analysis allows for realistic models to be parameterised and ensure they are better aligned with empirical findings. Nonetheless, many fascinating questions remain unanswered that will require greater focus on the consequences of changes in variance (rather than just the mean). Particularly interesting questions include: How does a reduction in physiological rate variance change energy flow across tropic levels within communities? What are the biochemical, cellular and physiological mechanisms that underly reduced variance in physiological rates? Are reductions in variance in one trait associated with reductions in correlated traits, or do some traits increase while others decrease? How much of the reduction in variance is driven by lower levels of genetic variance? Answers to these questions will require integrative approaches that combine empirical and theoretical work across multiple levels of biological organisation but will likely provide useful advances in understanding the full consequences that climate change will have on ectotherms across all major ecosystems.

## Funding
This work was partially supported by an Australian Research Council (ARC) Discovery Early Career Research Award (DE150101774) and ARC Future Fellowship (FT220100276) to DWAN.
